---
title: "Embedding flexdashboard in pkgdown site"
author: '@Rami_Krispin'
date: "9/6/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In the previous post, I demonstrated how to deploy a flexdashboard (or basically, any R Markdown format) in Github Pages. This post is focusing on embedding flexdashboard on pkgdown site. The pkgdown package is another great R Markdown format that enables us to quickly warp package information (documentation, vignettes, etc.) into a web format. 

### Motivation

Why would you like to embed flexdashboard in a pkgdown site? There few reasons that I can think about, for example:

* Provide a summary and analysis of the package datasets
* Demonstrate use case of the package functionality

My motivation for embedding flexdashboard into my pkgdown site was the first. A few weeks ago, while I was working on the coronavirus and covid19italy packages, I created a dashboard for each package with flexdashboard. As flexdashboard is another R Markdown format, I was looking for a method to render flexdashboard as one of the package vignettes. That did not work (or render) as I expected, but I found a workaround for that using the HTML iframe function.

### Embedding flexdashboard with iframe

The iframe function is an HTML for displaying a web page within a web page. To embed flexdashboard in pkgdown website, you will need:

* A deployed flexdashboard with access URL (see this post for instruction for deployment of flexdashboard on Github Pages)
* A rendered pkgdown

Once you have an end URL for the deployed flexdashboard, it is straightforward to deploy it by creating a dedicated vignette for embedding the dashboard as an article. The following example demonstrate step-by-step how to do it. We will start by creating a vignette, using the `use_vignette` function from the [usethis](https://usethis.r-lib.org/) package inside your package project:

``` r
usethis::use_vignette(name = "dashboard", title = "Your Dashboard Name")
```

This will create a R Markdown file `dashboard.Rmd` file inside the `vignettes` folder:

``` Rmarkdown
    ---
    title: "Your Dashboard Name"
    output: rmarkdown::html_vignette
    vignette: >
      %\VignetteIndexEntry{Your Dashboard Name}
      %\VignetteEngine{knitr::rmarkdown}
      %\VignetteEncoding{UTF-8}
    ---
    
    ```{r, include = FALSE}
    knitr::opts_chunk$set(
      collapse = TRUE,
      comment = "#>"
    )
    ```
    
    
  
```


The following example, from the covid19italy package, demonstrated the embedding of the Covid19 Italy Dashboard via a vignette:

`dashboard.Rmd`

``` Rmarkdown
    ---
    title: "Covid19 Italy Dashboard"
    output: rmarkdown::html_vignette
    vignette: >
      %\VignetteIndexEntry{intro}
      %\VignetteEngine{knitr::rmarkdown}
      %\VignetteEncoding{UTF-8}
    ---
    
    <style>
    .dashboard {
      position: fixed;
      left: 0;
      top: 50px;
      bottom: 0;
      right: 0;
    }
    .dashboard iframe {
      width: 100%;
      height: 100%;
      border: none;
    }
    </style>
    
    <div class="dashboard"> 
    <iframe src="https://ramikrispin.github.io/italy_dash/">
    </iframe>
    </div>
```